Community data modelling
========================================================


```{r}
library(ggplot2)
library(vegan)
library(wesanderson)

## Read in Community matrix and format it 

# Read in csv for Community Matrix (not normalized)
class = read.csv("~/Documents/AllAntibiotic_Class_notNormalized_antibioticClass.csv")
dim(class)
names(class)[1:10]
# Grab just numeric columns
classNumeric = class[,-1*c(1:7)]
# Eliminate columns that are all zeros
classNumeric=classNumeric[,colSums(classNumeric) !=0]
# Rebind string and numeric data into one dataframe
classDat = data.frame(class[,c(1:7)],classNumeric)
```


```{r}
## Read in genus data and format it (normalized)
genus = read.csv("~/Documents/AllAntibiotic_Genus.csv")
dim(genus)
names(genus)[1:10]
# Take only the numeric fields
genusNum = genus[,-1*c(1:6)]
# Get rid of fields that are 0
genusNum=genusNum[,colSums(genusNum) !=0]
# Scale everything to 1000 reads and round 
genusScale=apply(genusNum,1:2, function(x) 1e+03 * x)
genusScale = floor(genusScale)
# Rebind numeric and string data into one data frame
genusDat = data.frame(genus[,c(1:6)],genusScale)


```


```{r}
## Plot alpha diversity

# Plot alpha diversity for Genus
divG = data.frame(specnumber(genusScale))
colnames(divG) = "richness"
shannonG = data.frame(diversity(genusScale,index="shannon"))
colnames(shannonG) = "shannon"
type = genus[,2]
type=factor(type,levels(type)[c(4,2,1,5,3)])
shannonGplot=cbind(shannonG,type)
richnessGplot = data.frame(divG,type)

# Plot shannon diversity
ggplot(shannonGplot,aes(x=type,y=shannon,fill=type))+geom_boxplot()+ggtitle("Alpha diversity at genus level") + scale_fill_manual(values = wes.palette(5,"Darjeeling"))

# Do a t-test on just week and month
weekMonthIdx = shannonGplot$type == "week" | shannonGplot$type == "month"
weekMonth = shannonGplot[weekMonthIdx,]

t.test(weekMonth$shannon~weekMonth$type)

# Do a t-test on just week and multipleyear
weekMyearIdx = shannonGplot$type == "week" | shannonGplot$type == "multipleyear"
weekMyear = shannonGplot[weekMyearIdx,]

t.test(weekMyear$shannon~weekMyear$type)

# Plot richness
ggplot(richnessGplot,aes(x=type,y=richness,fill=type))+geom_boxplot()+ggtitle("Alpha diversity at genus level") + xlab("") + ylab("richness") + scale_fill_manual(values = wes.palette(5,"Darjeeling"))

# Do a t-test on just week and month
weekMonthIdx = richnessGplot$type == "week" | richnessGplot$type == "month"
weekMonth = richnessGplot[weekMonthIdx,]

t.test(weekMonth$richness~weekMonth$type)

# Do a t-test on just week and myear
weekMyearIdx = richnessGplot$type == "week" | richnessGplot$type == "multipleyear"
weekMyear = richnessGplot[weekMyearIdx,]

t.test(weekMyear$richness~weekMyear$type)

```
There is no significant difference in shannon's diversity between the week group and month group as well as wekk and multiple year group.

There is a significant difference in observed richness between the week group and the month group 
```{r}
# plot alpha diversity for class
divC = data.frame(specnumber(classNumeric))
colnames(divC) = "richness"
shannonC = data.frame(diversity(classNumeric,index="shannon"))
colnames(shannonC) = "shannon"
shannonCplot=cbind(shannonC,type)
richnessCplot = data.frame(divC,type)

# Plot shannon diversity for class
ggplot(shannonCplot,aes(x=type,y=shannon,fill=type))+geom_boxplot()+ggtitle("Alpha diversity at class level")+scale_fill_manual(values = wes.palette(5,"Darjeeling"))

# t-test to see if week is different than month
weekMonthIdx = shannonCplot$type == "week" | shannonCplot$type == "month"
weekMonth = shannonCplot[weekMonthIdx,]

t.test(weekMonth$shannon~weekMonth$type)

# t-test to see if week is different than multipleyear
weekMyearIdx = shannonCplot$type == "week" | shannonCplot$type == "multipleyear"
weekMyear = shannonCplot[weekMyearIdx,]

t.test(weekMyear$shannon~weekMyear$type)

ggplot(richnessCplot,aes(x=type,y=richness,fill=type))+geom_boxplot()+ggtitle("Alpha diversity at class level") + xlab("") + ylab("richness")+scale_fill_manual(values = wes.palette(5,"Darjeeling"))

# Do a t-test on just week and month
weekMonthIdx = richnessCplot$type == "week" | richnessCplot$type == "month"
weekMonth = richnessCplot[weekMonthIdx,]

t.test(weekMonth$richness~weekMonth$type)

# Do a t-test on just week and myear
weekMyearIdx = richnessCplot$type == "week" | richnessCplot$type == "multipleyear"
weekMyear = richnessCplot[weekMyearIdx,]

t.test(weekMyear$richness~weekMyear$type)

```
No significant differences at the class level between the week and month or multipleyear
```{r}
## Filter data set 

## Class level
# Select only samples who have taken antibiotics in the past week 
weekIDX =  class$Type == "week" 
week = class[weekIDX,]
dim(week)
# There are 38 samples that are week

# Select only samples who have taken antibiotics in the past month
monthIdx = class$Type == "month"
month = class[monthIdx,]
dim(month)
# THere are 59 samples that are month so we want to subsample to have equal size groups
set.seed(1)
monthSampIdx = sample(59,38)
monthSamp = month[monthSampIdx,]
dim(monthSamp)

# Combine into one dataset called recent of just samples from the past week and month
recentClass = rbind(monthSamp,week)
rowSums(recentClass[,-1*c(1:7)])
row.names(recentClass)
recentClass = recentClass[-40,]

## Genus level
weekIDX =  genusDat$Type == "week" 
week = genusDat[weekIDX,]
dim(week)
# There are 38 samples that are week

# Select only samples who have taken antibiotics in the past month
monthIdx = genusDat$Type == "multipleyear"
month = genusDat[monthIdx,]
dim(month)
# THere are 59 samples that are month so we want to subsample to have equal size groups
set.seed(1)
monthSampIdx = sample(1471,38)
monthSamp = month[monthSampIdx,]
dim(monthSamp)

# Combine into one dataset called recent of just samples from the past week and month
recentGenus = rbind(monthSamp,week)


```

```{r}
# Run random forests to see which factors separate these two groups
library(randomForest)

## Class level

# Get rid of extraneous columns
recentClassnum = recentClass[,-c(1:7)]

# Get rid of columns with all zeros
recentClassnum = recentClassnum[,colSums(recentClassnum) !=0]

# Normalize counts
classScale=t(apply(recentClassnum,1, function(x) 1e+03 * x/sum(x)))


# Add in factor of interest - in this case type
Type = recentClass$Type[drop=TRUE]
recentClassDat = data.frame(Type,recentClassnum)



# Run random forests
class.recent.rf <- randomForest(Type ~ ., data = recentClassDat, importance = TRUE, proximity = TRUE)
                                
print(class.recent.rf)

class.recent.mds <- cmdscale(1 - class.recent.rf$proximity, eig = TRUE)
varImpPlot(class.recent.rf,main="")

```
The results from this random forest show that Bacteroidia, Erysipelotrichi, Fusobacteria, Gammaproteobacteria, clostridia, and Bacilli are the most important factors for accuracy in classification. 

We will now split our samples into train and test groups and use a support vector machine with this small set of factors to try to correctly classify our samples


```{r}
## Use SVM on training set to train model and then validate with test model
library(e1071)

# Select only the columns listed as important from random forests
colIdx = colnames(recentClassDat) %in% c("X.Bacteroidia.", "X.Erysipelotrichi.","X.Fusobacteria_class.","X.Gammaproteobacteria.","X.Clostridia.","X.Bacilli.")
  type=recentClassDat[,1]
  svmCtrim = data.frame(type,recentClassDat[,colIdx])
  dim(svmCtrim)

# Number of repetitions
B=500
err=rep(0,B)

set.seed(3)

# Resample the test and train groups B times
for(b in 1:B){

  # separate train and test groups using a 60/40 split
  train=sample(nrow(svmCtrim),nrow(svmCtrim)*.6) 
  trainset=svmCtrim[train,]
  testset=svmCtrim[-train,]

  # Fit the SVM
  svmfit = svm(type~.,data=trainset,kernel="polynomial",cost=100,scale=TRUE)
  # Predict with the testset
  svm.pred <- predict(svmfit,testset[,-1])
  # Calculate the error rate from the confusion table
  errorTab = table(predict=svm.pred,truth=testset[,1])
  err[b] = (errorTab[2]+errorTab[3])/sum(errorTab)
}

# Mean error across all repetitions
mean(err)


# Lets plot the means of these classes between week and month
ggplot(svmCtrim, aes(x=type,y=X.Bacteroidia.)) + geom_boxplot()
ggplot(svmCtrim, aes(x=type,y=X.Clostridia.)) + geom_boxplot()


```


```{r}
# Lets try modelling with genus

# Get rid of extraneous columns
recentGenusnum = recentGenus[,-1*c(1:6)]

# Get rid of columns with all zeros
recentGenusnum = recentGenusnum[,colSums(recentGenusnum) !=0]
dim(recentGenusnum)

# Add in factor of interest - in this case type
Type = recentGenus$Type[drop=TRUE]
recentGenusDat = data.frame(Type,recentGenusnum)

# Run random forests
genus.recent.rf <- randomForest(Type ~ ., data = recentGenusDat, importance = TRUE, proximity = TRUE)

print(genus.recent.rf)

genus.recent.mds <- cmdscale(1 - genus.recent.rf$proximity, eig = TRUE)
varImpPlot(genus.recent.rf,main="",cex=.9)
```
 lets use Bacillus, Blautia, 313,323,604, and Allstipes


```{r}
## SVM 
library(e1071)

# Select only the columns listed as important from random forests
colIdx = colnames(recentGenusDat)== "X.958" | colnames(recentGenusDat)=="X.313" | colnames(recentGenusDat)== "X.1079" | colnames(recentGenusDat)== "X.Akkermansia."| colnames(recentGenusDat)== "X.505"| colnames(recentGenusDat)== "X.Acidaminococcus"
type=recentGenusDat[,1]
svmGtrim = data.frame(type,recentGenusDat[,colIdx])
dim(svmGtrim)

# Number of bootstraps
B=500
err=rep(0,B)

set.seed(3)

for(b in 1:B){

# separate train and test groups using a 60/40 split
train=sample(nrow(svmGtrim),nrow(svmGtrim)*.6) 
trainset=svmGtrim[train,]
testset=svmGtrim[-train,]

# Let's try svm
svmfit = svm(type~.,data=trainset,kernel="polynomial",cost=100,scale=TRUE)
svm.pred <- predict(svmfit,testset[,-1])
errorTab = table(predict=svm.pred,truth=testset[,1])
err[b] = (errorTab[2]+errorTab[3])/sum(errorTab)
}

mean(err)


# Lets plot the means of these genera between week and month
ggplot(svmGtrim, aes(x=type,y=X.Blautia.)) + geom_boxplot()
ggplot(svmGtrim, aes(x=type,y=X.323)) + geom_boxplot()
```

```{r}

# Lets look at categories of antibiotics

# Seperate Class 1's
class1Idx = recentClass$Class1 == 1
class1 = recentClass[class1Idx,]
class1$ClassOther = "Class1"

#Seperate Class 2's
class2Idx = recentClass$Class2 != 0
class2 =  recentClass[class2Idx,]
class2$ClassOther = "Class2"

# Rebind Class 1 and 2
classClass = rbind(class1,class2)

# Repeat for genus

# Add in factor of interest - in this case type
recentGenusData=data.frame(recentGenus[,c(1:6)],recentGenusnum)
anticlass = recentGenusData$ClassOther[drop=TRUE]
recentGenusDat = data.frame(anticlass,recentGenusnum)

# Seperate Class 1's
class1Idx = recentGenusData$Class1 == 1
class1 = recentGenusData[class1Idx,]
class1$ClassOther = "Class1"

#Seperate Class 2's
class2Idx = recentGenusData$Class2 != 0
class2 =  recentGenusData[class2Idx,]
class2$ClassOther = "Class2"

# Rebind Class 1 and 2
classGenus = rbind(class1,class2)

```

```{r}

library(ape)

# Hierarchical clustering by time since antibiotic use
recentGenusnum = recentGenusnum[-40,]
bdist = vegdist(recentGenusnum,method="bray")

hclust.bc = hclust(bdist,method="average")

colsType=c(rep("red",38),rep("blue",37))
plot(as.phylo(hclust.bc),direction="downwards",tip.color=colsType,cex=.7,main="Hierarchical clustering of samples at genus level")
legend("topleft","week","month")


# Hierarchical clustering by class of antibiotic
recentclassClassnum = classClass[,-1*c(1:7)]
bdist = vegdist(recentclassClassnum,method="bray")

# Dendogram 
hclust.bc = hclust(bdist,method="average")
colsClass=c(rep("red",30),rep("blue",25))
plot(as.phylo(hclust.bc),direction="downwards",tip.color=cols,cex=.8,main="Hierarchical clustering of samples at genus level")
legend("topright",c("week","month"))


# Bray curtis distance looking at class of antibiotic


comm.mds=metaMDS(recentclassClassnum,dist="bray")

mds.fig = ordiplot(comm.mds,type="none")
points(mds.fig,"sites",pch=19,col="blue",select=classClass$ClassOther=="Class1")
points(mds.fig,"sites",pch=19,col="red",select=classClass$ClassOther=="Class2")

adonis(bdist~ClassOther,data=classClass)
```

```{r}
# Redo random forests

classGenus$ClassOther <- as.factor(classGenus$ClassOther)
classGenusdat = classGenus[,-c(1:5)]

# Run random forests
class.genus.rf <- randomForest(ClassOther ~ ., data = classGenusdat, importance = TRUE, proximity = TRUE)

print(class.genus.rf)

class.class.mds <- cmdscale(1 - class.genus.rf$proximity, eig = TRUE)
varImpPlot(class.genus.rf)

```

```{r}
## Redo SVM with antibiotic classes

# Select only the columns listed as important from random forests
colIdx = colnames(classClassdat)== "X.Methanobacteria." | colnames(classClassdat)=="X.Clostridia." | colnames(classClassdat)== "X.Sphingobacteria." | colnames(classClassdat)== "X.Gammaproteobacteria."| colnames(classClassdat)== "X.Flavobacteria."| colnames(classClassdat) == "X.Fusobacteria_class."| colnames(classClassdat) == "X.Bacteroidia."
type=classClassdat[,2]
classClassNum = classClassdat[,-c(1,2)]
svmCtrim = data.frame(type,classClassdat[,colIdx])
dim(svmCtrim)

# Number of bootstraps
B=500
err=rep(0,B)

set.seed(3)

for(b in 1:B){

# separate train and test groups using a 60/40 split
train=sample(nrow(svmCtrim),nrow(svmCtrim)*.7) 
trainset=svmCtrim[train,]
testset=svmCtrim[-train,]

# Let's try svm
svmfit = svm(type~.,data=trainset,kernel="linear",cost=100,scale=TRUE)
svm.pred <- predict(svmfit,testset[,-1])
errorTab = table(predict=svm.pred,truth=testset[,1])
err[b] = (errorTab[2]+errorTab[3])/sum(errorTab)
}

mean(err)
```

```{r}

# Redo random forests at genus level with antibiotic class

row.names(re)

# Run random forests
class.class.rf <- randomForest(ClassOther ~ ., data = classClassdat, importance = TRUE, proximity = TRUE)

print(class.class.rf)

class.class.mds <- cmdscale(1 - class.class.rf$proximity, eig = TRUE)
varImpPlot(class.class.rf)
## Redo SVM with antibiotic genera

# Select only the columns listed as important from random forests
colIdx = colnames(classGenusdat)== "X.958" | colnames(classGenusdat)=="X.880" | colnames(classGenusdat)== "X.90" | colnames(classGenusdat)== "X.675"
type=classGenusdat[,1]
classGenusNum = classGenusdat[,-1]
svmCtrim = data.frame(type,classGenusdat[,colIdx])
dim(svmCtrim)

# Number of bootstraps
B=500
err=rep(0,B)

set.seed(3)

for(b in 1:B){

# separate train and test groups using a 60/40 split
train=sample(nrow(svmCtrim),nrow(svmCtrim)*.7) 
trainset=svmCtrim[train,]
testset=svmCtrim[-train,]

# Let's try svm
svmfit = svm(type~.,data=trainset,kernel="linear",cost=100,scale=TRUE)
svm.pred <- predict(svmfit,testset[,-1])
errorTab = table(predict=svm.pred,truth=testset[,1])
err[b] = (errorTab[2]+errorTab[3])/sum(errorTab)
}

mean(err)

# Now with genera